\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-26 analysis/manuscript.tex
Birgits suggestions

For over two decades, hybrid open access journal publishing, which makes
some articles openly available while others remain behind a paywall, has
been discussed as a means for transitioning the subscription system to
full open access (Prosser, 2003). The idea was that when journals
increasingly publish open access articles, they could reduce revenues
from subscriptions, while libraries and funders could change their
funding models and shift expenditures from subscription to open access.
However, initial approaches, mainly based on publication fees, also
called article processing charges (APCs), did not contribute
substantially to a large open access uptake. In 2009, Springer reported
open access to 1\% of articles in hybrid journals (Dallmeier-Tiessen et
al., 2010). Other studies also recorded a low uptake. In 2011, only
1-2\% of articles were open access (Björk, 2012), growing to around 4\%
between 2011 and 2013 (Laakso \& Björk, 2016).

With the introduction of central funding mechanisms for publication fees
in some European countries since 2012, an increase in hybrid open access
could be observed (Björk, 2017; Huang et al., 2020; Jubb et al., 2017;
Piwowar et al., 2018). For example, studying university output,
Robinson-Garcia et al. (2020) estimated a median uptake of 7.1\% in the
period 2014-2017. In particular, British (17\%), Austrian (15\%) and
Dutch (13\%) universities stand out. However, this shift in funding
policy towards hybrid open access also added to the overall cost of
publishing, which includes subscription spending and the administrative
efforts required to handle payments (Pinfield et al., 2016). Moreover,
established large commercial publishers, which already dominated the
publishing market (Larivière et al., 2015), disproportionately benefited
from hybrid open access funding in comparison to full open access
publishers (Butler et al., 2023; Jahn \& Tullney, 2016; Shu \&
Larivière, 2023).

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and open access publishing (Mittermaier, 2015), as well as to increase
publisher-provided immediate open access (Björk \& Solomon, 2014;
Schimmer et al., 2015). These considerations resulted into
transformative agreements\footnote{In this paper I use the term
  ``transformative agreement'', addressing also offsetting,
  read-and-publish or publish-and-read deals, and other variants
  (Borrego et al., 2021; Hinchliffe, 2019). Although the term is
  critised as misleading and not useful to describe the different types
  of open access agreements between library consortia and commercial
  publishers (Babini et al., 2022), it is widely used in policy
  discussions and in the research literature.}, which cover a broad
range of contracts between library consortia and publishers from
mid-2010s onwards where institutional spending for subscriptions and
open access publishing are considered together (Borrego et al., 2021;
Hinchliffe, 2019). Transformative agreements seek to control costs while
allowing a transitional phase for publishing more open access articles.
Similar to big deals, transformative agreements mainly bundle hybrid and
subscription-only journals from commercial publishers, but aim at a
higher degree of transparency than previous big deals, where contracts
including payments were confidential (Bergstrom et al., 2014).

The introduction of transformative agreements aligns with funding policy
changes, such as the decision made by the cOAlition S, a consortium of
national funders including the European Commission, to no longer provide
financial support for individual publication fees when publishing in
hybrid journals. According ot its Plan S launched in 2018, the cOAlition
S funders only accept hybrid open access through transformative
agreements ``during a transition period that should be as short as
possible'' (Schiltz, 2018). Specifically, they agreed to support hybrid
open access only through transformative agreements from 2021, until the
end of 2024. Notably, the German Research Foundation (DFG), despite not
being part of cOAlition S, has also extended its financial support for
hybrid open access through transformative agreements (Mittermaier,
2021). Previously, the DFG only provided funding for full open access
journals (Jahn \& Tullney, 2016).

By the end of 2023, many transformative agreements were implemented, but
interim outcomes are mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both full open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilization of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuate market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality (Klebel \& Ross-Hellauer, 2023), because transformative
agreements focus on pay to publish open access mainly targets
institutions from high-income countries, furthering a questionable
journal prestige culture (Babini et al., 2022). Besides, an
editorial-board resignation raised concerns that publishers' desire to
maximize journal publication volume ``without regard to quality'' is a
consequence of transformative agreements (Rasmussen, 2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
Association of Swedish Higher Education Institutions (Sveriges
universitets- och högskoleförbund, SUHF) recommended to only support
agreements for publishing in full open access journals \footnote{\url{https://www.su.se/english/news/open-access-need-to-move-away-from-transformative-agreements-1.683787}}.
Likewise, most cOAlition S funders will discontinue financial support
for transformative agreements by the end of 2024 (Liverpool, 2023). The
consortium also removed most hybrid journals from its Transformative
Journal program in 2023 due to publishers' failure to meet self-defined
open access growth targets (Brainard, 2023). In contrast, the German
DEAL consortium announced a five-year transformative agreement with
Elsevier starting in 2024, while also renewing its contracts with
Springer Nature and Wiley until the end of 2028. At the same time, the
Colombia Consortium signed the first transformative agreements in Latin
America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this uptake can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018), while large-scale studies
relied on self-reported agreement data (Moskovkin et al., 2022), or used
APC pricing lists (Shu \& Larivière, 2023). Particularly, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements. Likewise, article-level open access invoicing
data is only partly available (Jahn et al., 2022).

The present study aims to address these limitations by combining
multiple openly available data sources to determine the open access
uptake in hybrid journals, while distinguishing between open access
through transformative agreements and other means. With this novel and
open approach, this first large-scale analysis aims to answer the
following questions:

@@

This study combines data from multiple publicly available data sources
as shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from the cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identification of hybrid journals by excluding full
open access journals, Crossref served as the primary data source for
article-level metadata including Creative Commons (CC) license
information to indicate open access availability on publisher websites.
To determine open access articles published through transformative
agreements, first author affiliations from OpenAlex (Priem et al., 2022)
were subsequently linked to eligible institutions according to the
transformative agreement data. In the following, these steps are
described in more detail.

@@

A limitation of using the Journal Checker Tool and its underlying
publicly available data dump to study the development of transformative
agreements over time is that expired transformative agreements are
constantly removed. To address this, four different snapshots were
safeguarded and combined for this study: self-archived versions from
July 2021, July 2022, and May 2023, as well as the most current dump
downloaded on 11 December 2023. This ensured that transformative
agreements, which ended from 2021 onwards, were included, representing
the majority of transformative agreements. Overall, the four combined
Transformative Agreement Data dumps used in this study contained 729 out
of 869 agreements listed in the ESAC registry by December 2023.

The Transformative Agreement Data dumps link agreements to journals
represented by journal names and ISSN. After mapping ISSN variants to
the corresponding linking ISSN (ISSN-L) as provided by the ISSN
International Centre, journals were associated to publishers using the
ESAC ID, a unique identifier for transformative agreements in the ESAC
Transformative Agreement Registry. Furthermore, journal subjects
according to the All Science Journal Classification code (ASJC) were
added from the Scopus journal source list as of August 2023.

Because transformative agreements can include both full open access and
hybrid journals, the data were complemented with information about a
journal's open access status using multiple sources: the Directory of
Open Access Journals (DOAJ) downloaded on 12 December 2023\footnote{\url{https://doaj.org/csv}},
OpenAlex (November 2023) and the the Bielefeld list of GOLD OA journals
(Bruns et al., 2022). As shown in Figure \ref{fig:method_fig}A,
combining different data sources considerably extended the journal
matching. In total, 3,439 full open access journals were excluded based
on ISSN matching. The overlap between the three data sources was 72\%.
The Gold OA journals dataset alone added 176 journals, while the DOAJ
comprised 10 full open access journals not listed in either of the other
two sources. These full open access journals were mostly launched in
2022.

@@

After identifying hybrid journals included in transformative agreements,
article metadata was retrieved from the Crossref November 2023 database
snapshot for the five-year period 2018 to 2022 according to the issued
date, representing the earliest known publication date. Because Crossref
metadata lacked information to distinguish between original research
articles including review and other types of journal content, which are
often not covered by transformative agreements (Borrego et al., 2021),
only articles published in regular issues indicated by non-numeric
pagination were included. Furthermore, an expanded version of
Unpaywall's paratext recognition approach was applied to exclude
non-scholarly journal content such as table of contents.

Open access articles in hybrid journals were identified through Creative
Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version were
not considered. Crossref was used for open access identification because
transformative agreements workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017).

Comparing Crossref license coverage with OpenAlex, which re-uses open
access evidence from Unpaywall, a widely used open access discovery
service that also parses journal websites for open content licenses
(Piwowar et al., 2018), highlighted ongoing challenges to identify
hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
with CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The biggest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers only differ slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable difference
could be furthermore observed among some publishers that presumably did
not provide CC license information to Crossref including AIP Publishing,
American Physiological Society, Emerald and the Royal Society. Crossref
license metadata was more complete with regard to articles from the
publisher Wiley and American Chemical Society. Finally, inconsistent
open access status information in previous OpenAlex versions was
observed (Jahn et al., 2023). After reporting it to OpenAlex, fixing
this issue was still ongoing according to the release notes, which might
also explain part of the discrepancy.

After retrieving article metadata, the publication volume including open
access was calculated per journal. To improve the identification of
hybrid journals, journals with an open access proportion above 95\% were
excluded. This further step allowed to remove additional 241 full open
access journals.

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements, because they are
considered to be responsible to arrange open access publication (Borrego
et al., 2021; Geschuhn \& Stone, 2017; Schimmer et al., 2015). Here,
country and institutional affiliations were retrieved from OpenAlex.
However, because of low coverage in OpenAlex, this study focused on
first authors and their affiliations instead. First authors typically
contribute most to a paper and are often considered lead author research
papers (Larivière et al., 2016). Related studies therefore assumed first
authors as a proxy to measure to open access payments and the impact of
transformative agreement (Haucap et al., 2021; Shu \& Larivière, 2023;
Zhang et al., 2022). Overall, around 90\% of studied articles had first
author affiliation metadata in OpenAlex, whereas the coverage of
articles with corresponding author information was around 54\%.

To assess the impact of transformative agreements to hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with first author
affiliations recorded by OpenAlex using the ROR-ID. The matching also
took into account the duration of an agreement according to the ESAC
registry. Upon inspection, Transformative Agreement Data did not cover
associated institutions comprehensively like university hospitals or
institutes of large research organisations like the Max Planck Society.
To improve the matching, Transformative Agreement Data was automatically
enriched with associated organisations using OpenAlex's institution
entity data.

In total, the compiled data set consists of 8,922,146 articles published
in 12,857 hybrid journals between 2018 and 2022 (see Figure
\ref{fig:method_fig}B). Hybrid journals in transformative agreements
represented 40\% of total global output over the same time period
according to Crossref, while full open access journals recorded 35\%.

@@

\caption{Initial data characteristics. (A) Full open access journals included in transformative agreements by evidence source Directory of Open Access Journals (DOAJ), OpenAlex and the Bielefeld GOLD OA list. (B) Number of articles in Crossref by journal types. The blue bars show the overall article volume of hybrid journals in transformative agreements, which were initially included in the study, in comaprision with full open access journals according to OpenAlex. The remainder represents closed access journals not covered by transformative agreements.}\label{fig:method_fig}
\end{figure}

@@

Throughout this mostly automated data gathering and analysis process,
tools from the Tidyverse (Wickham et al., 2019) for the R programming
language (R Core Team, 2020) were used. The resulting data is openly
available through an R data package, hoaddata, version 0.2.91 Following
Marwick et al. (2018), hoaddata contains not only the datasets used in
the data analysis. It also includes code used to compile the data by
connecting it the cloud-based Google Big Query data warehouse, where the
big scholarly data from Crossref, OpenAlex and Unpaywall were imported.
To increase the computational reproducibility, the data aggregation
through hoaddata was automatically carried out using GitHub Actions, a
continuous integration service.

@@

Between 2018 and 2022, a total of 11,189 out of 12,857 hybrid journals
in transformative agreements published at least one open access article
under a Creative Commons license. During this period, these hybrid
journals provided open access to 742,369 out of 8,146,958 articles,
representing a five-year open access proportion of 9.1\%. Authors who
could make use of transformative agreements at the time of publication
contributed 328,957 open access articles to the total.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the
proportion of open access articles in hybrid journals, comparing the
overall open access uptake and the impact of transformative agreements
on this trend. Over the five-years period from 2018 to 2022, open access
increased from 4.3\% (n = 65,486) to 15\% (n = 249,511). At the same
time, the total article volume of the investigated journals grew from
1,528,051 in 2018 to 1,676,928 in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022, contributing 58\% of the total open access article
volume in 2022. However, there was also a notable growth in open access
provision through other means, presumably publication fees being not
invoiced through transformative agreements, which increased from 4.1\%
(n = 62,625) in 2018 to 6.3\% (n = 105,896).

Figure \ref{fig:results_overview}C depicts the substantial variations
among the hybrid journals included in transformative agreements in terms
of open access uptake. Although the median generally follows the trend
shown in Figure \ref{fig:results_overview}A, the farther stretch of
upper quartiles and whiskers over the years illustrates that an
increasing number of journals published an above-average proportion of
open access articles. In 2022, 25\% of hybrid journals (n = 2,576) had
an open access uptake of 29\%, and 6.6\% of journals (n = 744) provided
the majority of their articles under a Creative Commons license in the
same year. These journals were, on average, smaller (M = 75, SD = 186)
than those with an open access share below 50\% (M = 164, SD = 347).

When comparing the impact of open access trough transformative
agreements across journals, it shows that for many journals these
agreements substantially contributed to the growth of open access over
the years (Figure \ref{fig:results_overview}D). Despite the rise in
transformative agreements, it is worth noting that other means of
publishing open access remained common across the investiagted hybrid
journals. In total, 9,153 journals published open access articles from
authors affiliated with institutions without transformative agreements
in place, while 8,780 journals published at least one open access
article through a transformative agreement in the same year.

@@

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley accounted for 49\% of total article volume
published (see Table \ref{tab:publisher_league_table}). Together, they
published 500,878 or 66\% of open access articles in hybrid journals.
Elsevier, Springer Nature, and Wiley made 243,891 articles open access
in hybrid journals through transformative agreements, resulting in an
even larger market share of 74\%.

@@

However, there are notable differences among the three big publishers.
Although Elsevier published the largest volume of articles (n =
2,770,826, 34\%), it recorded a comparable low number of open access
articles, including those that can be associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open acccess), leading to
higher open access market shares (23\% Springer Nature resp. 23\%
Wiley). This difference between Elsevier on the one hand and Springer
Nature and Wiley on the other can be attributed to transformative
agreements, as the latter made the majority of their open access
articles available through such deals (Springer Nature 57\% resp. Wiley
55\%).

>>>>

For over two decades, hybrid open access journal publishing, which makes
some articles openly available while others remain behind a paywall, has
been discussed as a means for transitioning the subscription system to
full open access (Prosser, 2003). The idea was that when journals
increasingly publish open access articles, they could reduce revenues
from subscriptions, while libraries and funders could change their
funding models and shift expenditures from subscription to open access.
However, initial approaches, mainly based on publication fees, also
called article processing charges (APCs), did not contribute
substantially to a large open access uptake. In 2009, Springer reported
open access to 1\% of articles in hybrid journals (Dallmeier-Tiessen et
al., 2010). Other studies have also recorded low uptake. In 2011, only
1-2\% of articles were open access (Björk, 2012), increasing to around
4\% between 2011 and 2013 (Laakso \& Björk, 2016).

With the introduction of central funding mechanisms for publication fees
in some European countries since 2012, an substantial increase in hybrid
open access has been observed (Björk, 2017; Huang et al., 2020; Jubb et
al., 2017; Piwowar et al., 2018). For example, studying university
output, Robinson-Garcia et al. (2020) estimated a median uptake of 7.1\%
in the period 2014-2017. In particular, British (17\%), Austrian (15\%)
and Dutch (13\%) universities stood out. However, this shift in funding
policy towards hybrid open access also added to the overall cost of
publishing, which includes subscription spending and the administrative
efforts required to handle payments (Pinfield et al., 2016). Moreover,
large commercial publishers, which already dominated the publishing
market (Larivière et al., 2015), disproportionately benefited from
hybrid open access funding in comparison to full open access publishers
(Butler et al., 2023; Jahn \& Tullney, 2016; Shu \& Larivière, 2023).

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and open access publishing (Mittermaier, 2015), as well as to increase
publisher-provided immediate open access (Björk \& Solomon, 2014;
Schimmer et al., 2015). These considerations resulted in transformative
agreements\footnote{In this paper I use the term ``transformative
  agreement'', addressing also offsetting, read-and-publish or
  publish-and-read deals, and other variants (Borrego et al., 2021;
  Hinchliffe, 2019). Although the term is critised as misleading and not
  useful to describe the different types of open access agreements
  between library consortia and commercial publishers (Babini et al.,
  2022), it is widely used in policy discussions and in the research
  literature.}, which cover a broad range of contracts between library
consortia and publishers from the mid-2010s onwards, where institutional
spending for subscriptions and open access publishing are considered
together (Borrego et al., 2021; Hinchliffe, 2019). Transformative
agreements seek to control costs while allowing a transitional phase for
publishing more open access articles. Similar to big deals,
transformative agreements mainly bundle hybrid and subscription-only
journals from commercial publishers, but aim at a higher degree of
transparency than previous big deals, where contracts including payments
were confidential (Bergstrom et al., 2014).

The introduction of transformative agreements aligns with funding policy
changes, such as the decision made by cOAlition S, a consortium of
funders and research organisations including the European Commission, to
no longer provide financial support for individual publication fees when
publishing in hybrid journals. According to its Plan S launched in 2018,
cOAlition S members only accept hybrid open access through
transformative agreements ``during a transition period that should be as
short as possible'' (Schiltz, 2018). Specifically, they agreed to
support hybrid open access only through transformative agreements from
2021, until the end of 2024. Notably, despite not being part of
cOAlition S, the German Research Foundation (DFG), has also extended its
financial support for hybrid open access through transformative
agreements (Mittermaier, 2021). Previously, the DFG only provided
funding for fully open access journals (Jahn \& Tullney, 2016).

By the end of 2023, many transformative agreements had been implemented,
but the interim outcomes were mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both fully open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilisation of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements has contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuates market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality (Klebel \& Ross-Hellauer, 2023), because transformative
agreements' focus on pay to publish mainly targets institutions from
high-income countries, furthering a questionable journal prestige
culture (Babini et al., 2022). Besides, an editorial-board resignation
raised concerns that transformative agreements might encourage
publishers to maximize journal publication volume ``without regard to
quality'' (Rasmussen, 2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
Association of Swedish Higher Education Institutions (Sveriges
universitets- och högskoleförbund, SUHF) recommended only supporting
agreements for publishing in fully open access journals \footnote{\url{https://www.su.se/english/news/open-access-need-to-move-away-from-transformative-agreements-1.683787}}.
Likewise, most cOAlition S funders will discontinue financial support
for transformative agreements by the end of 2024 (Liverpool, 2023). The
consortium also removed the majority of hybrid journals from its
Transformative Journal program in 2023 due to publishers' failure to
meet self-defined open access growth targets (Brainard, 2023). By
contrast, the German DEAL consortium announced a five-year
transformative agreement with Elsevier starting in 2024 and also renewed
its contracts with Springer Nature and Wiley until the end of 2028.
Similarly, the Colombia Consortium signed the first transformative
agreements in Latin America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018), while large-scale studies
relied on self-reported agreement data (Moskovkin et al., 2022), or used
APC pricing lists (Shu \& Larivière, 2023). In particular, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements. Likewise, article-level open access invoicing
data are only partly available (Jahn et al., 2022).

The present study aims to address these limitations by combining
multiple openly available data sources to determine open access uptake
in hybrid journals, while distinguishing between open access through
transformative agreements and other means. With this novel and open
approach, this first large-scale analysis will answer the following
questions:

@@

This study combines data from multiple publicly available sources, as
shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identifying hybrid journals by excluding fully open
access journals, Crossref served as the primary data source for
article-level metadata including Creative Commons (CC) license
information to indicate open access availability on publisher websites.
To determine open access articles published through transformative
agreements, first author affiliations from OpenAlex (Priem et al., 2022)
were subsequently linked to eligible institutions according to the
transformative agreement data. In the following, the steps are described
in detail.

@@

A limitation of using the Journal Checker Tool and its underlying
publicly available data dump to study the development of transformative
agreements over time is that expired transformative agreements are
constantly removed. To address this, four different snapshots were
safeguarded and combined for this study: self-archived versions from
July 2021, July 2022, and May 2023, as well as the most current dump
downloaded on 11 December 2023. This ensured that transformative
agreements, which ended from 2021 onwards, were included, representing
the majority of transformative agreements. Overall, the four combined
Transformative Agreement Data dumps used in this study contained 729 of
869 agreements listed in the ESAC registry by December 2023.

The Transformative Agreement Data dumps link agreements to journals
represented by journal names and the ISSN. After mapping ISSN variants
to the corresponding linking ISSN (ISSN-L), as provided by the ISSN
International Centre, journals were associated with publishers using the
ESAC ID, a unique identifier for transformative agreements in the ESAC
Transformative Agreement Registry. Furthermore, journal subjects
according to the All Science Journal Classification code (ASJC) were
added from the Scopus journal source list as of August 2023.

Because transformative agreements can include both fully open access and
hybrid journals, the data were complemented with information about a
journal's open access status using multiple sources: the Directory of
Open Access Journals (DOAJ) downloaded on 12 December 2023\footnote{\url{https://doaj.org/csv}},
OpenAlex (November 2023) and the Bielefeld list of GOLD OA journals
(Bruns et al., 2022). As shown in Figure \ref{fig:method_fig}A,
combining different data sources considerably extended journal matching.
In total, 3,439 full open access journals were excluded based on ISSN
matching. The overlap between the three data sources was 72\%. The Gold
OA journals dataset alone added 176 journals, while the DOAJ comprised
10 fully open access journals not listed in either of the other two
sources. These fully open access journals were launched in 2022.

@@

After identifying hybrid journals included in transformative agreements,
article metadata were retrieved from the Crossref November 2023 database
snapshot for the five-year period from 2018 to 2022, according to the
issued date, representing the earliest known publication date. Because
Crossref metadata lacked information to distinguish between original
research articles, including reviews, and other types of journal
content, which are often not covered by transformative agreements
(Borrego et al., 2021), only articles published in regular issues
indicated by non-numeric pagination were included. Furthermore, an
expanded version of Unpaywall's paratext recognition approach was
applied to exclude non-scholarly journal content such as table of
contents.

Open access articles in hybrid journals were identified using the
Creative Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version was not
considered. Crossref was used for open access identification because
transformative agreement workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017).

Comparing Crossref license coverage with OpenAlex, which re-uses open
access evidence from Unpaywall, a widely used open access discovery
service that also parses journal websites for open content licenses
(Piwowar et al., 2018), highlighted ongoing challenges in identifying
hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
under CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The largest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers differ only slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable differences
could be observed among some publishers that presumably did not provide
CC license information to Crossref, including AIP Publishing, the
American Physiological Society, Emerald, and the Royal Society. Crossref
license metadata was more complete with regard to the articles published
by Wiley and the American Chemical Society. Finally, inconsistent open
access status information in previous OpenAlex versions was observed
(Jahn et al., 2023). After reporting this to OpenAlex, fixing this issue
was still ongoing according to the release notes, which might also
explain part of the discrepancy.

After retrieving the article metadata, the publication volume, including
open access, was calculated for each journal. To improve the
identification of hybrid journals, journals with an open access
proportion above 95\% were excluded. This step allowed removing
additional 241 fully open access journals.

Affiliation metadata about corresponding authors are crucial for the
planning and evaluation of transformative agreements because they are
considered responsible for arranging open access publication (Borrego et
al., 2021; Geschuhn \& Stone, 2017; Schimmer et al., 2015). For this
study, country and institutional affiliations were retrieved from
OpenAlex. However, because corresponding author affiliation was only
available for 54\% of the investigated articles in OpenAlex, this study
focused on first authors and their affiliations instead; approximately
90\% of the investigated articles had first author affiliation metadata
in OpenAlex. First authors typically contribute the most to a paper and
are often considered being lead authors (Larivière et al., 2016).
Related studies also assumed first authors as a proxy to measure open
access payments and the impact of transformative agreements (Haucap et
al., 2021; Shu \& Larivière, 2023; Zhang et al., 2022).

To assess the impact of transformative agreements on hybrid open access,
participating institutions from the Transformative Agreement Data dump,
which were crowd-sourced from the agreements and consortia that
successfully negotiated an agreement, were matched with the first author
affiliations recorded by OpenAlex using the ROR-ID. Matching also
considered the duration of agreements according to the ESAC registry.
Upon inspection, Transformative Agreement Data did not comprehensively
cover associated institutions, such as university hospitals or
institutes of large research organisations such as the Max Planck
Society. To improve the matching, Transformative Agreement Data were
automatically enriched with associated organisations using OpenAlex's
institution entity data.

The so compiled data set consists of 8,922,146 articles published in
12,857 hybrid journals between 2018 and 2022 (see Figure
\ref{fig:method_fig}B). Hybrid journals in transformative agreements
represented 40\% of the total global output over the same time period,
according to Crossref, while fully open access journals recorded 35\%.

@@

\caption{Initial data characteristics. (A) Full open access journals included in transformative agreements by evidence source Directory of Open Access Journals (DOAJ), OpenAlex and the Bielefeld GOLD OA list. (B) Number of articles in Crossref by journal types. The blue bars show the overall article volume of hybrid journals in transformative agreements, which were initially included in the study, in comparision with full open access journals according to OpenAlex. The remainder represents subscription-based journals not covered by transformative agreements.}\label{fig:method_fig}
\end{figure}

@@

Throughout this mostly automated data gathering and analysis process,
Tidyverse tools (Wickham et al., 2019) for the R programming language (R
Core Team, 2020) were used. The resulting data are openly available
through an R data package, hoaddata, version 0.2.91 Following Marwick et
al. (2018), hoaddata contains not only the datasets used in the data
analysis. It also includes code used to compile the data by connecting
it to the cloud-based Google Big Query data warehouse, where scholarly
big data from Crossref, OpenAlex and Unpaywall were imported. To
increase computational reproducibility, data aggregation through
hoaddata was automatically carried out using GitHub Actions, a
continuous integration service.

@@

Between 2018 and 2022, 11,189 out of 12,857 hybrid journals in
transformative agreements published at least one open access article
under a Creative Commons license. During this period, these hybrid
journals provided open access to 742,369 of the 8,146,958 articles,
representing a five-year open access proportion of 9.1\%. Authors who
could make use of transformative agreements at the time of publication
contributed 328,957 open access articles.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the
proportion of open access articles in hybrid journals, comparing the
overall open access uptake and the impact of transformative agreements
on this trend. Over the five-years period from 2018 to 2022, open access
increased from 4.3\% (n = 65,486) to 15\% (n = 249,511). Simultaneously,
the total article volume of the investigated journals increased from
1,528,051 in 2018 to 1,676,928 in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022, contributing 58\% of the total open access article
volume in 2022. However, there was also a notable increase in open
access provision through other means, presumably publication fees being
not invoiced through transformative agreements, which increased from
4.1\% (n = 62,625) in 2018 to 6.3\% (n = 105,896).

Figure \ref{fig:results_overview}C depicts the substantial variations
among the hybrid journals included in transformative agreements in terms
of open access uptake. Although the median generally follows the trend
shown in Figure \ref{fig:results_overview}A, the farther stretch of
upper quartiles and whiskers over the years illustrates that an
increasing number of journals published an above-average proportion of
open access articles. In 2022, 25\% of hybrid journals (n = 2,576) had
an open access uptake of 29\%, and 6.6\% of journals (n = 744) provided
the majority of their articles under a Creative Commons license in the
same year. On average, these journals were smaller (M = 75, SD = 186)
than those with an open access share below 50\% (M = 164, SD = 347).

When comparing the impact of open access through transformative
agreements across journals, it shows that for many journals, these
agreements substantially contributed to the growth of open access over
the years (Figure \ref{fig:results_overview}D). Despite the rise in
transformative agreements, it is worth noting that other means of
publishing open access remained common across the investigated hybrid
journals. In total, 9,153 journals published open access articles from
authors affiliated with institutions without transformative agreements
in place, while 8,780 journals published at least one open access
article through a transformative agreement in the same year.

@@

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of the total article
volume published (see Table \ref{tab:publisher_league_table}). Together,
they published 500,878 or 66\% of the open access articles in hybrid
journals. Elsevier, Springer Nature, and Wiley made 243,891 articles
open access in hybrid journals through transformative agreements,
resulting in an even larger market share of 74\%.

@@

However, there are notable differences between the three large
publishers. Although Elsevier published the largest volume of articles
(n = 2,770,826, 34\%), it recorded a comparably low number of open
access articles, including those that are associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open access), leading to higher
open access market share (23\% Springer Nature resp. 23\% Wiley). This
difference between Elsevier on the one hand and Springer Nature and
Wiley on the other can be attributed to transformative agreements, as
the latter made the majority of their open access articles available
through such deals (Springer Nature 57\% resp. Wiley 55\%).
